# This is a heading (large text)

## Subheading (smaller)
### Subsubheading (even smalller) --> then you knit to show this 

Some text. some _italics- or some *italics*.  Some__bold__ with double underscores or some **bold**

my list: (create bullets below)

* item
* another item
  * subitem
  
My ordered list:

1. item 1
2. item 2
3. item 3 

words from a wise man:
> Here are some words
> more words  - these all get smooshed into one line unless you use two double space

Inline code. The `paste()` function is great.  above the squiggle under esc 

col 1  |  Col 2    
-------|---------
x1     | x2
x2     | x4


#Git or github for markdown

# Title it 
# Hit "Knit" button
# name the file


#a .Rmd file has 3 major types of content: the header, the Markdown content and the R code chunks

# YAML3 format title 
# ---
# title: "EXAMPLE_TITLE"
# author: "YOUR_NAME"
# date: "2/01/2018"
# output: html_document
# ---

## Second Level Header
#This is just plain markdown that can contain **bold** or _italics_
# 
```{r}
# R code chunk in an R Markdown file
some_variable <- 100
```

# write code in separate doc and then use the source() function 

# Write normal **markdown** out here, then create a code block:

  ```{r}
# Execute R code in here
course_number <- 201
```

# Back to writing _markdown_ out here.

```{r options_example, echo = FALSE, message = TRUE)
# A code chunk named "options_example", with argument `echo` assigned FALSE
# and argument `message` assigned TRUE

# Would execute R code in here
```

echo - indicates you want the R code itself in the document 
True = display or False = do not display

message - indicates if you want any messages generated by code to be displayed

include - indicates if any results of code should be in output of report 


```{r list_example, echo = FALSE}


markdown_list <- "
- Lions
- Tigers
- Bears
- Oh mys
"
```

```{r pasted_list_example, echo = FALSE}
# Create a vector of animals
animals <- c("Lions", "Tigers", "Bears", "Oh mys")

# Paste `-` in front of each animal and join the items together with
# newlines between
markdown_list <- paste("-", animals, collapse = "\n")
```

`r markdown_list`

#dateframe rendering in Markdown 

```{r kable_example, echo = FALSE}
library("knitr") # make sure you load the package (once per document)

# Make a data frame
letters <- c("a", "b", "c", "d")
numbers <- 1:4
df <- data.frame(letters = letters, numbers = numbers)

# "Return" the table to render it
kable(df)
```

`r markdown_list`

#rendering plots - test by clicking green play button icon above each chunk

```{r plot_example, echo = FALSE}
library("ggplot2") # make sure you load the package (once per document)

# Plot of college education vs. poverty rates in the Midwest
ggplot(data = midwest) +
  geom_point(
    mapping = aes(x = percollege, y = percadultpoverty, color = state)
  ) +
  scale_color_brewer(palette = "Set3")
```