---
title: "Markdown2"
author: "Jenny Skytta"
date: "03/15/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(knitr)
```

# This is an analysis of breakfast cereals

```{r}
dataframe <- read.csv("balkdjf;akldfj", stringsAsFactors = FALSE)

sugars_by_mfr <- cereal %>% 
  group_by(mfr) %>% 
  summarize(mean_sugars = mean(sugars), .groups = 'drop')

highest_sugars <- cereal %>%  
  filter(sugars == max(sugars)) %>% 
  pull(name)

lowestt_sugars <- cereal %>%  
  filter(sugars == min(sugars)) %>% 
  pull(name)

top_five <- cereal %>% 
  select(name, sugars) %>% 
  arrange(-sugars) %>% 
  top_n(5, sugars)
```

## Highest/lowest sugars
We found that the cereals with the highest sugars were `r highest_sugars`. Meanwhile the cereals with the lowest amount of sugars was `r lowest_sugars`. This is a table with the top five most sugary cereals:

```{r echo = FALSE}
kable(top_five)
```

## Rating vs. Sugars
This is a plot that shows the relationship between sugars and rating of the cereal.  We see tehre is a negative correlation. 

```{r echo = FALSE}
plot1 <- ggplot(data = cereal)  +
  geom_point(mapping = aes(x = rating, y = sugars, color = mfr), alpha = 0.5)
  xlab("Rating") +
  ylab("Sugars(g)") +
  ggitle("Rating vs. sugars for cerals") +
    scale_fill_brewer(palette = "Dark2", name = "Manufacturer")

show(plot1)

```

