---
title: 'IMT 573: Module 3 Lab'
subtitle: 'Advanced Visualization'
author: "Jenny Skytta"
date: 'Due: April 18, 2022'
output: 
  pdf_document:
    includes:
      in_header: "preamble.tex"
---

<!-- This syntax can be used to add comments to a markdown file that are ignored during knitting process. -->
##### Collaborators: _Independent work_

### Objectives

As we continue our data science journey, we are gaining skills in working with data. This might be reflected in more efficient ways to manipulate and summarize data, both of which can be useful for creating more advanced visualizations of that data. To accomplish many of the visualization tasks in these exercises you will need to make use of newly acquired data manipulation skills!

### Instructions

Before beginning this assignment, please ensure you have access to R and RStudio; this can be on your own personal computer or on the IMT 573 R Studio Cloud. 

1. Open the `03_lab_advancedviz.Rmd` and save a copy to your local directory. Supply your solutions to the assignment by editing `03_lab_advancedviz.Rmd`. 

2. First, replace the "YOUR NAME HERE" text in the `author:` field with your own full name. Any collaborators must be listed on the top of your assignment. 

3. Be sure to include well-documented (e.g. commented) code chucks, figures, and clearly written text chunk explanations as necessary. Any figures should be clearly labeled and appropriately referenced within the text. Be sure that each visualization adds value to your written explanation; avoid redundancy -- you do no need four different visualizations of the same pattern.

4.  Collaboration on problem sets is fun and useful, and I encourage it, but each student must turn in an individual write-up in their own words as well as code/work that is their own.  Regardless of whether you work with others, what you turn in must be your own work; this includes code and interpretation of results. The names of all collaborators must be listed on each assignment. Do not copy-and-paste from other students' responses or code.

5. All materials and resources that you use (with the exception of lecture slides) must be appropriately referenced within your assignment.  

6. When you have completed the assignment and have **checked** that your code both runs in the Console and knits correctly when you click `Knit`. When the PDF report is generated rename the knitted PDF file to `lab3_YourLastName_YourFirstName.pdf`, and submit the PDF file on Canvas.

### Setup

In this lab you will need, at minimum, the following R packages.

```{r Setup, message=FALSE}
# Load standard libraries
library(tidyverse)
```

The data we will use in this lab comes from the Million Song Dataset. The Million Song Dataset is a collaboration between the Echo Nest and LabROSA, a laboratory working towards intelligent machine listening. The project was also funded in part by the National Science Foundation of America (NSF) to provide a large data set to evaluate research related to algorithms and information retrieval. 

[http://millionsongdataset.com/](You can find out ---
title: 'IMT 573: Module 3 Lab'
subtitle: 'Advanced Visualization'
author: "YOUR NAME HERE"
date: 'Due: July 8, 2021'
output: 
  pdf_document:
    includes:
      in_header: "preamble.tex"
---

<!-- This syntax can be used to add comments to a markdown file that are ignored during knitting process. -->
##### Collaborators: Independent work

List collaborators here.

### Objectives

As we continue our data science journey, we are gaining skills in working with data. This might be reflected in more efficient ways to manipulate and summarize data, both of which can be useful for creating more advanced visualizations of that data. To accomplish many of the visualization tasks in these exercises you will need to make use of newly acquired data manipulation skills!

### Instructions

Before beginning this assignment, please ensure you have access to R and RStudio; this can be on your own personal computer or on the IMT 573 R Studio Cloud. 

1. Open the `03_lab_advancedviz.Rmd` and save a copy to your local directory. Supply your solutions to the assignment by editing `03_lab_advancedviz.Rmd`. 

2. First, replace the "YOUR NAME HERE" text in the `author:` field with your own full name. Any collaborators must be listed on the top of your assignment. 

3. Be sure to include well-documented (e.g. commented) code chucks, figures, and clearly written text chunk explanations as necessary. Any figures should be clearly labeled and appropriately referenced within the text. Be sure that each visualization adds value to your written explanation; avoid redundancy -- you do no need four different visualizations of the same pattern.

4.  Collaboration on problem sets is fun and useful, and I encourage it, but each student must turn in an individual write-up in their own words as well as code/work that is their own.  Regardless of whether you work with others, what you turn in must be your own work; this includes code and interpretation of results. The names of all collaborators must be listed on each assignment. Do not copy-and-paste from other students' responses or code.

5. All materials and resources that you use (with the exception of lecture slides) must be appropriately referenced within your assignment.  

6. When you have completed the assignment and have **checked** that your code both runs in the Console and knits correctly when you click `Knit`. When the PDF report is generated rename the knitted PDF file to `lab3_YourLastName_YourFirstName.pdf`, and submit the PDF file on Canvas.

### Setup

In this lab you will need, at minimum, the following R packages.

```{r Setup, message=FALSE}
# Load standard libraries
library(tidyverse)
install.packages("maps")
library(maps)

```

The data we will use in this lab comes from the Million Song Dataset. The Million Song Dataset is a collaboration between the Echo Nest and LabROSA, a laboratory working towards intelligent machine listening. The project was also funded in part by the National Science Foundation of America (NSF) to provide a large data set to evaluate research related to algorithms and information retrieval. 

[http://millionsongdataset.com/](You can find out more information about the dataset here.)

Thierry Bertin-Mahieux, Daniel P.W. Ellis, Brian Whitman, and Paul Lamere. The Million Song Dataset. In Proceedings of the 12th International Society for Music Information Retrieval Conference (ISMIR 2011), 2011.

We will use a subset of this data created by Ryan Whitcomb, \tt{rwhit94@vt.edu}, which contains data on 10,000 songs.  The data contains standard information about the songs such as artist name, title, and year released. Additionally, the data contains more advanced information; for example, the length of the song, how many musical bars long the song is, and how long the fade in to the song was.

```{r read in data, message=FALSE}
# Load music data
#music_data <- read_csv("Downloads/music.csv")
```

### Problem 1: Inspection 

First, inspect the data. You can use functions such as `glimpse`, `head`, `tail`, etc. to help you get a sense of what is contained in the data. 
```{r, message=FALSE, warning=FALSE}

glimpse(music_data) #condensed view of dataframe
head(music_data) # first few observations of dataframe
tail(music_data) #last few observations of dataframe
dim(music_data) # dimensions of dataframe 

```


### Problem 2: Pose a Question

Propose a question to guide your analysis. For example, you might ask if the average hotness scores of songs change over time? Or perhaps, what is the relationship between song duration and tempo? You can use one of these questions or develop your own. State which question you want to answer.

What is song hotttnesss you ask? According to the dataset description, it is a measure of the song's popularity, when downloaded (in December 2010). And measured on a scale of 0 to 1.	

## Questions 

- _What are the top genres of music that are the most "hott" (frequently listed) ?_
- _Where was the location of the artist's with the oldest recordings?_

```{r}
# create new tibble 
hott_genre <- music_data %>% 
  rename(genre = artist.terms, 
         song_hotness = song.hotttnesss, 
         artist_hotness = artist.hotttnesss) %>% # artist.terms rename to "genre"
  group_by(genre) %>%  # group_by genre
  summarise(frequency = n(), # summarise n() of genre
            song_hotness = max(song_hotness), 
            artist_hotness = max(artist_hotness)) %>% 
  arrange(desc(song_hotness)) 
   # arrange by artist.hotttnessss and song.hotttnesss
   
#View(hott_genre)   

# create new tibble
oldest_song_location <- music_data %>% 
  group_by(artist.name) %>% # group_by artist 
  arrange(desc(-song.year)) %>%  # arrange by descending year
  filter(artist.latitude > 1 & artist.longitude > 1 & song.year > 1) %>% #filter to remove 0 values
  select(artist.name, song.year, artist.latitude, artist.longitude)
         # select artist.latitude, artist.longitude 
# head(data, 5) to get the top 5
# 
```


### Problem 3: Visualization

Create two visualizations to help gain insight into your question. Be sure to explain the visuals you create and what you take away from them.  
```{r}
hott_genre_5 <- head(hott_genre, 5)
#View(hott_genre_5)



oldest_song_loc_5 <- head(oldest_song_location, 10)
#View(oldest_song_loc_5)
world <- map_data("world")

  ggplot(data = world) +
    geom_sf() +
    coord_sf(xlim = c(0, 75), ylim = c(35, 70), expand = FALSE) +
  geom_point(
    data = oldest_song_loc_5,
    aes(artist.longitude, artist.latitude, color = artist.name),
    alpha = 2.7
  ) +
  theme_void() +
    labs(
    title = "Oldest Song Artist's Geographic World locations",
    color = "Artists"
    )

```



### Citations

Thierry Bertin-Mahieux, Daniel P.W. Ellis, Brian Whitman, and Paul Lamere. 
The Million Song Dataset. In Proceedings of the 12th International Society
for Music Information Retrieval Conference (ISMIR 2011), 2011.

https://datavizpyr.com/how-to-make-world-map-with-ggplot2-in-r/ more information about the dataset here.)

Thierry Bertin-Mahieux, Daniel P.W. Ellis, Brian Whitman, and Paul Lamere. The Million Song Dataset. In Proceedings of the 12th International Society for Music Information Retrieval Conference (ISMIR 2011), 2011.

We will use a subset of this data created by Ryan Whitcomb, \tt{rwhit94@vt.edu}, which contains data on 10,000 songs.  The data contains standard information about the songs such as artist name, title, and year released. Additionally, the data contains more advanced information; for example, the length of the song, how many musical bars long the song is, and how long the fade in to the song was.

```{r read in data, message=FALSE}
# Load music data
music_data <- read_csv("Downloads/music.csv")
View(music_data)
```

### Problem 1: Inspection 

First, inspect the data. You can use functions such as `glimpse`, `head`, `tail`, etc. to help you get a sense of what is contained in the data. 

### Problem 2: Pose a Question

Propose a question to guide your analysis. For example, you might ask if the average hotness scores of songs change over time? Or perhaps, what is the relationship between song duration and tempo? You can use one of these questions or develop your own. State which question you want to answer.

What is song hotttnesss you ask? According to the dataset description, it is a measure of the song's popularity, when downloaded (in December 2010). And measured on a scale of 0 to 1.	

### Problem 3: Visualization

Create two visualizations to help gain insight into your question. Be sure to explain the visuals you create and what you take away from them. 
